A quantitative Mac Diarmid's inequality for geometrically ergodic Markov chains

### 2020-01-07 Note_exponential_moment/MarkovCon.tex
updat

@@

\section{Main result}\label{sec:Main}
The main result of this paper  is the following quantitative version of Mac Diarmid's inequality for $V-$ergodic Markov chains under certain drift condition.
%\begin{theorem}\label{thm:ConcMarkQuant}
 %Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
 %\[
%\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
% \]
%where $\beta$ is given by
 %\[
 %\beta=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
 %\]
%\end{theorem}
\begin{theorem}\label{thm:ConcMarkVErgodic}
 Assume {\bf H1}, {\bf H2} and {\bf H3} holds with function $V(x)$. Let $\cC = \{V(x) < d\}$, $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
 \[
\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
 \]
where $\beta$ is given by
 \[
 \beta=\frac{(1-r\vee \lambda^{1/4})^2}{16Ld}\bigg(-\frac5{\log \lambda}+4MLd\bigg)^{-1}\eqsp.
 \]
\end{theorem}

@@

\begin{lemma}
\label{lem:drift_prod}
 Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{gamma}$. Then it holds
$$
Q^{n,p}V \leq \lambda^{\Gamma_{p,n}}V + C\frac{\lambda^{\Gamma_{1,p}} - \lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$ 
\end{lemma}
\begin{proof} For notation simplicity let us consider only $p=1$. Then by induction
$$
Q^{n}V \leq \lambda^{\Gamma_{1,n}}V + C \sum\limits_{k=1}^{n}\gamma_k \lambda^{\Gamma_{1,k-1}} \leq \lambda^{\Gamma_{1,n}}V + C \int\limits_{0}^{\Gamma_{1,n}}\lambda^{x}\,dx \leq
$$
$$
\leq \lambda^{\Gamma_{1,n}}V + C\frac{1-\lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$
\end{proof}
Let us check that ULA kernel \ref{} satisfies the drift condition \label{eq:drift}. We will need additional assumption
\par
[{\bf C2}] There exists such constants $K_1 > 0$ and $m > 0$, such that for any $x \notin B(0,K_1)$ it holds
$$
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
$$
Another popular condition on the potential is strong convexity outside some euclidean ball:
\par
[{\bf C3}] There exist constants $K_1 > 0$ and $m > 0$ such that for any $x \notin B(0,K_1)$ and $y \in \mathbb{R}^d$, it holds $\langle D^2U(x)y,y\rangle \geq m\|y\|^2$. Moreover, there exists $M \geq 0$ such that for any $x \in \mathbb{R}^d$, $\|D^3U(x)\| \leq M$.
\par
Yet it is possible to show that assumption {\bf C3} and {\bf C1} imply {\bf C2} (see \cite[Lemma~14]{brosse:2018:diffusion}), so in the sequel we are going to work with the condition {\bf C2}. It allows us to prove the following result

\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2}. Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1$.
\end{lemma}

@@

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf C1} and {\bf C2}
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$
$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2 + 2\|\nabla U(0)\|^2
$$
Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) + 2\gamma^2\|\nabla U(0)\|^2 \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(d + 2\gamma \|\nabla U(0)\|^2)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2 + 2\gamma^2\|\nabla U(0)\|^2$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d + 2\gamma \|\nabla U(0)\|^2\right) \leq 
$$

@@

>>>>

\begin{equation}
\label{eq:exp_moment_hit}

@@

\section{Main result}\label{sec:Main}
The main result of this paper  is the following quantitative version of Mac Diarmid's inequality for $V-$ergodic Markov chains under certain drift condition.
%\begin{theorem}\label{thm:ConcMarkQuant}
 %Assume {\bf H1}, {\bf H2}, {\bf H3}. Let $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
 %\[
%\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
% \]
%where $\beta$ is given by
 %\[
 %\beta=\frac{(1-r\vee u^{-1/4})^2}{16L}\bigg(\frac5{\log u}+4ML\bigg)^{-1}\eqsp.
 %\]
%\end{theorem}
\begin{theorem}\label{thm:ConcMarkVErgodic}
 Assume {\bf H1}, {\bf H2} and {\bf H3} holds with drift function $V(x)$. Let $M$ be a constant from \ref{eq:exp_moment_hit}, $\cC = \{V(x) < d\}$, $n\geqslant 1$, $c\in \R^n$ and  $f\in \functionboundeddiff{\sX^n}{c}$. Then, for all $x\in \cC$ and $t > 0$,
 \[
\Pbb_x\big(f( \chunk[0]{X}[n-1])-\Ebb_x[f(\chunk[0]{X}[n-1])]>t\big)\leqslant \exp\bigg(-\frac{\beta t^2}{\|c\|^2}\bigg)\eqsp,
 \]
where $\beta$ is given by
 \[
 \beta=\frac{(1-r\vee \lambda^{1/4})^2}{16Ld}\bigg(-\frac5{\log \lambda}+4MLd\bigg)^{-1}\eqsp.
 \]
\end{theorem}

@@

\begin{lemma}
\label{lem:drift_prod}
 Assume that $P_{\gamma}$ satisfies condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma}$. Then it holds
$$
Q^{n,p}V \leq \lambda^{\Gamma_{p,n}}V + C\frac{\lambda^{\Gamma_{1,p}} - \lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$ 
\end{lemma}
\begin{proof} For notation simplicity let us consider only $p=1$. Then by induction
$$
Q^{n}V \leq \lambda^{\Gamma_{1,n}}V + C \sum\limits_{k=1}^{n}\gamma_k \lambda^{\Gamma_{1,k-1}} \leq \lambda^{\Gamma_{1,n}}V + C \int\limits_{0}^{\Gamma_{1,n}}\lambda^{x}\,dx \leq
$$
$$
\leq \lambda^{\Gamma_{1,n}}V + C\frac{1-\lambda^{\Gamma_{1,n}}}{\log{\frac{1}{\lambda}}}
$$
\end{proof}
Let us check that ULA kernel \ref{} satisfies the drift condition \label{eq:drift}. We will need additional assumption
\par
[{\bf C2}] There exists such constants $K_1 > 0$ and $m > 0$, such that for any $x \notin B(0,K_1)$ it holds
$$
\langle \nabla U(x), x \rangle \geq \frac{m}{2}\|x\|^2
$$
Another popular condition on the potential is strong convexity outside some euclidean ball:
\par
[{\bf C3}] There exist constants $K_1 > 0$ and $m > 0$ such that for any $x \notin B(0,K_1)$ and $y \in \mathbb{R}^d$, it holds $\langle D^2U(x)y,y\rangle \geq m\|y\|^2$. Moreover, there exists $M \geq 0$ such that for any $x \in \mathbb{R}^d$, $\|D^3U(x)\| \leq M$.
\par
Yet it is possible to show that assumption {\bf C3} and {\bf C1} imply {\bf C2} (see \cite[Lemma~14]{brosse:2018:diffusion}), so in the sequel we are going to work with the condition {\bf C2}. It allows us to prove the following result

\begin{lemma} Assume that the potential $U(x)$ satisfies conditions {\bf C1} and {\bf C2} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel \ref{} satisfies drift condition \ref{eq:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{-\left(\frac{m}{2}\right)}$, $C = 3K_1^2 + 2d + C_1$ for some absolute constant $C_1 > 0$.
\end{lemma}

@@

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that $\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2$ and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf C1} and {\bf C2}
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$
$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $P_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + 2\gamma d
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
P_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^d\right) \leq 
$$

@@

Now the geometric ergodicity result readily follows from \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018}, namely, the following lemma holds:

\begin{lemma} Let the potential $U(x), x \in \mathbb{R}^D$ satisfy conditions {\bf C1} and {\bf C2}. Then for $0 < \gamma < \overline{\gamma}=\frac{m}{4L^2}$, for any $x \in X$ it holds
$$
d_V(\delta_xQ^n_\gamma,\pi_\gamma) \leq C\rho^n\left(V(x) + \pi_\gamma(V)\right)
$$ 
with $V(x) = 1 + \|x\|^2$ and constants
$$
b = \gamma(3K_1^2 + 2D + C_1); \quad \overline{b} = b\exp^{-\frac{m\gamma}{2}} + d; \quad \varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right); \quad
$$

$$
C = \left(1 + \exp^{-\frac{m\gamma}{2}}\right)\left(1+\frac{\overline{b}}{(1-\varepsilon)(1-\exp^{-\frac{m\gamma}{2}} - \frac{2b}{1+d})}\right);
$$

$$
\rho = \exp^{-\gamma\left(\frac{m}{2} - 2\frac{3K_1^2 + 2D + C_1}{d}\right)\frac{\log{(1-\varepsilon)}}{\log{(1-\varepsilon)} + \log{(\exp^{-\frac{m\gamma}{2}} + \frac{2b}{d+1})}}}
$$
for some $C_1 > 0$, which does not depend on $d, \gamma$ and $n$.
\end{lemma}

\begin{proof} Note that the condition {\bf C1} implies that the Markov kernel $Q^n_\gamma$ satisfies $(1,\varepsilon)$-Doeblin condition with $\varepsilon = 2\Phi\left(-\frac{\sqrt{d}(1+L\gamma)}{2\sqrt{\gamma}}\right)$. Together with drift condition {\bf C2} it allows to apply \cite[Theorem~19.4.1]{douc:moulines:priouret:soulier:2018} with appropriate constants. 
\end{proof}
